### 2024-02-27 doc/15yrsChange.Rmd
annual mortality data, figures 4 & 5
Minor revisions to introduction

Nuisance pests and pathogens have had a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al. 2016, Fischer et al. 2013, Herms and McCullough 2014, Fei et al. 2014).  The presence of nonindigenous forest insects and diseases, which have been historically introduced to new ecosystems through intercontinental trade, will continue to proliferate with increased globalization and human movement (Fischer et al. 2013, Aukema et al. 2010). Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy speciation and structure.  Once abundant in temperate deciduous forests throughout North America, the American Chestnut (*Castanea dentata*) and the Ash genera (*Fraxinus spp.*) have been decimated by pests and pathogens, with stand-clearing disturbances occuring within a few years after initial infestation (Anderson 2020, Klooster et al. 2013, Need a source for Ash abundance and length of time between infestation and mortality for Chestnuts). These previously dominant canopy species are now confined to the understory as non-reproductive individuals (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).  In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with risk for regional extirpation (Ellison et al. 2018).  The loss of mature canopy species to introduced nuisance species contributes to global change driven biomass losses, with estimates that nonindigenous pest and pathogens account for an additional 5.53 TgC per year (Fei et al. 2013).  A significant percentage of temperate forests remain at risk for future pest and pathogens (Fei et al. 2013).

>>>>

Nuisance pests and pathogens have had a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al. 2016, Fischer et al. 2013, Herms and McCullough 2014, Fei et al. 2014).  The presence of nonindigenous forest insects and diseases, which have been historically introduced to new ecosystems through intercontinental trade, will continue to proliferate with increased globalization and human movement (Fischer et al. 2013, Aukema et al. 2010, Brockerhoff et al. 2006). Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy speciation and structure.  Once abundant in temperate deciduous forests throughout North America, the American Chestnut (*Castanea dentata*) and the Ash genera (*Fraxinus spp.*) have been decimated by pests and pathogens, with stand-clearing disturbances occuring within a few years after initial infestation (Anderson 2020, Klooster et al. 2013, Need a source for Ash abundance and length of time between infestation and mortality for Chestnuts). These previously dominant overstorey species are now confined to the understory as non-reproductive seeds and saplings (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).  In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with considerable risk of regional extirpation (Ellison et al. 2018).  The loss of mature canopy species to nuisance species contributes to global change driven biomass losses, with biomass losses due to pest and pathogens comparable to those attributed to fire and forest removal (Fei et al. 2013)  Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al. 2013)
